---
title: "Reproducible Research: Peer Assessment 2"
output:
  html_document:
    keep_md: true
---

# Impact of Weather Events on Public Health and Economy

## Abstract

In this document, the impact of weather events on public health and economy across
the United State is evaluated.
The analysis is done based on the NOAA Storm Database.
First an exploratory analysis of the data is done,
to understand how to reduce the amount of available data
without compromising the expressiveness with respect to the addressed objectives.
From that, a data processing strategy is derived.
Eventually the impact of weather events on public health and economy are examined,
showing that tornados have the greatest impact on public health,
and floods on economy.

## Prerequisites

```{r load_libraries}
library(dplyr)
library(stringr)
```

## Raw Data

The raw dataset is downloaded from the URL provided with the Course Assignment: [StormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
Most of the columns of the datasets can be discared as they are not relevant for
understanding the relationship between weather events and poulation health and economy across USA.
Data with an unknown weather event have been removed.

```{r download_and_extract}
if (!dir.exists("tmp")) {
  dir.create("tmp")
}
raw_data_file_path <- file.path("tmp", "StormData.csv.bz2")
if (!file.exists(raw_data_file_path)) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
    raw_data_file_path
  )
}
```

```{r load_data, cache = TRUE}
raw_data <- as_tibble(read.csv(raw_data_file_path))
```

## Exploratory Analysis

### Data columns

The dataset contains a lot of columns, but only some of them seem to be relevant
for the present discussion.

```{r names}
names(raw_data)
```
Only the following variables seem to be relevant:

- `EVTYPE`: the type of weather event
- `FATALITIES`: The number of deaths caused by a weather event
- `INJURIES`: The number of injuries caused by a weather event
- `PROPDMG`: The estimated amount of damage to property incurred by the weather event
- `PROPDMGEXP`: The unit of the property damage measurement
- `CROPDMG`: The estimated amount of damage to crops incurred by the weather event
- `CROPDMGEXP`: The unit of the crop damage measurement

### Weather Event Types

There is a large number of different values in the `EVTYPE` column.
The values do not seem to be normalized:

 - some values seem to be invalid or not useful (e.g. `?`, `Summary August 10`),
 - some values are representing the same weather event (e.g. `TSTM WIND` and `THUNDERSTORM WINDS`).

> For details see `Appendix - List of Event Types`.

To avoid a normalization of the entire set of given event types, it makes sense to
take a look at the overall distribution of event types,
to find out if it is safe to focus on just a smaller subset of the events.

``` {r distribution_of_weather_events}
N <- dim(raw_data)[1]
by_evt_count <- raw_data %>%
  group_by(EVTYPE) %>%
  summarise(n = n()) %>%
  arrange(desc(n))
with(by_evt_count,
  plot(cumsum(n[1:10] / N), ylab = "Cumulative Sum of Events", xlab = "Weather Event Index")
)
```

**Distribution of Event Types**: The top ten events with highest frequency
  cover nearly 90% of the overall cases.

The top-ten event types contain some redundant values, i.e. different values
for the same weather event: `TSTM WIND`, `THUNDERSTORM WIND`, and `THUNDERSTORM WINDS`.
Thus, a normalization joining these values makes sense.

```{r top_ten_event_types}
by_evt_count[1:10, 'EVTYPE']
```

### Fatalities and Injuries

The given data for fatalities and injuries does not indicate any need for preprocessing.

```{r, fatalaties_and_injuries}
summary(raw_data[, c('FATALITIES', 'INJURIES')])
```

### Property Damage

Property damage is expressed via two different columns: `PROPDMG` and `PROPDMGEXP`

- `PROPDMG` provides a numeric value
- `PROPDMGEXP` defines how the value of `PROPDMG` has to be multiplied

For `PROPDMGEXP` the following values can be found:

```{r propdmgexp}
unique(raw_data$PROPDMGEXP)
```

The provided documentation does not explain how all of these values should
be interpreted.
The frequency of values indicates that it is safe to focus
only on the following values (assuming the common interpretation)

- `B` (billion),
- `M` (million),
- `K` (thousand),
- `H` (hundred)

```{r tbl_propdmgexp}
table(raw_data$PROPDMGEXP)
```
> Note: there are samples in lower and upper case.

### Crop Damage

Crop damage is expressed via two different columns: `CROPDMG` and `CROPDMGEXP`

- `CROPDMG` provides a numeric value
- `CROPDMGEXP` defines how the value of `CROPDMG` has to be multiplied

For `CROPDMGEXP` the following values can be found:

```{r cropdmgexp}
unique(raw_data$CROPDMGEXP)
```

Similar to `PROPDMGEXP` there are values which are difficult to interpret.

```{r tbl_cropdmgexp}
table(raw_data$CROPDMGEXP)
```

The table of frequencies indicates that it is admissible to focus on these values:

- `B` (billion),
- `M` (million),
- `K` (thousand),
- `H` (hundred)

> Note: similar to `PROPDMGEXP` there are values in lower and upper case

## Data Preprocessing

From the Exploratory Data Analysis the following actions are indicated:

- Selecting relevant columns
- Removing rows without EVTYPE
- Joining EVTYPE variants for "THUNDERSTORM WIND"
- Subsetting to the Top-Ten Events

```{r pre_processing}
data <- raw_data %>%
  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  filter(EVTYPE != "?") %>%
  mutate(EVTYPE = case_when(
    EVTYPE %in% c("TSTM WIND", "THUNDERSTORM WIND", "THUNDERSTORM WINDS") ~ "THUNDERSTORM WIND",
    TRUE ~ EVTYPE)
  )
data_by_event_count <- data %>% group_by(EVTYPE) %>% summarise(n = n()) %>% arrange(desc(n))
top_ten_events <- data_by_event_count[1:10,]$EVTYPE
top_ten_events_data = data %>% filter(EVTYPE %in% top_ten_events)
```

## Impact on Public Health

The top three weather events with respect to injuries and fatalities are
tornados, floods, and thunderstorm winds.
With respect to only fatalities they are tornados, flash floods, and lightings.

```{r impact_on_public_health}
  top_ten_events_data %>%
  group_by(EVTYPE) %>%
  summarise(total_injuries = sum(INJURIES),
            total_fatalities = sum(FATALITIES),
            total_casualities = sum(INJURIES + FATALITIES)) %>%
  arrange(desc(total_casualities))
```
**Impact on Public Health**: tornados are the source for most injuries as well as fatalities.

## Impact on Economy

The top three weather events with respect to economic consequences are floods, tornados, and hails.
With respect to just property damage they are floods, tornados, and flash floods.
With respect to just crop damage they are floods, hail, and flash floods.

```{r impact_on_economy}
  top_ten_events_data %>%
  mutate(
    property_damage = case_when(
      str_detect(PROPDMGEXP, regex('B', ignore_case = TRUE)) ~ 1000000000 * PROPDMG,
      str_detect(PROPDMGEXP, regex('M', ignore_case = TRUE)) ~ 1000000 * PROPDMG,
      str_detect(PROPDMGEXP, regex('K', ignore_case = TRUE)) ~ 1000 * PROPDMG,
      str_detect(PROPDMGEXP, regex('H', ignore_case = TRUE)) ~ 100 * PROPDMG,
      TRUE ~ 0
    ),
    crop_damage = case_when(
      str_detect(CROPDMGEXP, regex('B', ignore_case = TRUE)) ~ 1000000000 * CROPDMG,
      str_detect(CROPDMGEXP, regex('M', ignore_case = TRUE)) ~ 1000000 * CROPDMG,
      str_detect(CROPDMGEXP, regex('K', ignore_case = TRUE)) ~ 1000 * CROPDMG,
      str_detect(CROPDMGEXP, regex('H', ignore_case = TRUE)) ~ 100 * CROPDMG,
      TRUE ~ 0
    )
  ) %>%
  group_by(EVTYPE) %>%
  summarise(
    total_property_damage = sum(property_damage),
    total_crop_damage = sum(crop_damage),
    total_damage = sum(property_damage + crop_damage)
  ) %>%
  arrange(desc(total_damage)) %>%
  mutate(
    total_property_damage = paste(round(total_property_damage / 1000000000, digits = 2) , ' B'),
    total_crop_damage = paste(round(total_crop_damage / 1000000000, digits = 2) , ' B'),
    total_damage = paste(round(total_damage / 1000000000, digits = 2) , ' B'),
  )
```
**Economic Consequences**: floods are the source for most property and crop damages.

## Result

The impact of weather events on public health and economy across the United States have
been investigated. The analysis reveals that tornados have the greatest impact on public health,
while floods have the greatest impact on economy.

# Appendix

## List of Event Types

```{r event_types}
sort(unique(raw_data$EVTYPE))
```
